Differentiable Optics with {\Large $\partial$}Lux: I - Deep Calibration of Flat Field and Phase with autodiff

### 2022-11-17 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{spie}
% \documentclass[]{spie}
% \documentclass[twocolumn]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage{xspace}
%\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations

\renewcommand{\baselinestretch}{1.0} % Change to 1.65 for double spacing
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
% \usepackage{hyperref}
\usepackage{relsize}
\usepackage{acro}
\acsetup{
  single    = false,
  list/sort  = true,
  cite/group = true,
  cite/group/cmd = \cite,
  cite/group/pre = {},
}

@@

@@

\author[a]{Louis C. Desdoigts}
\author[b]{Benjamin J. S. Pope}
\author[a]{Peter G. Tuthill}

% Begin!
\authorinfo{Send correspondence to L. C. D.: \\E-mail: \href{mailto:louis.desdoigts@sydney.edu.au}{louis.desdoigts@sydney.edu.au}}

@@

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, by taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework \equinox \cite{kidger2021equinox}, with an optical system consisting of an \equinox module wrapping a stack of `layers' which are themselves \equinox layers applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

@@

\begin{enumerate}
    \item Wavefront initialisation
    \item Application of the pupil-support
    \item Normalisation of the wavefront
    \item Application of the pupil phase mask
    \item Application of the optical-aberrations
    \item Propagation to the focal plane
\end{enumerate}

@@

Next the wavefront encounters the binary phase plate diffractive pupil from Toliman \cite{tuthill2018}, with a half-wave total \ac{opd} between the binary regions. We use this as an information-preserving way to spread the \ac{psf} across a large number of pixels which encodes the PRF information for a large number of pixels simultaneously. Other conventional methods such as defocusing or using a diffuser \cite{Stefansson2017} can be employed but result in a larger degradation in the information carried by the \ac{psf} and therefore result in all round worse performance. An important consideration which is alleviated by the Toliman pupil is that focal-plane wavefront sensing suffers a sign ambiguity for all phase modes that are inversion-symmetric about the origin, and for unambiguous phase retrieval we therefore need an asymmetric or odd-mode symmetric pupil \cite{Martinache2013}. This pupil and corresponding aberrated \ac{psf} are shown in the left panel of Figure \ref{fig:optics}.

The wavefront then encounters low-order phase errors parameterised by coefficients on Zernike polynomials. We include the first 4 radial terms, ignoring global piston, tip and tilt for a total of 7 terms. Coefficients were randomly drawn from a normal distribution of mean 0 and standard deviation 20nm, for a total of \textcolor{red}{showyourwork variance} nm RMS total \ac{opd}. These aberrations applied to the pupil mask and resulting \ac{psf} are shown in the middle panel of Figure~\ref{fig:optics}. As can be seen from the \ac{psf} this is a large amount of optical distortion.

This is then propagated to the 1024x1024 pixel focal plane by a two-sided matrix Fourier transform \cite{Soummer2007,Martinache2020}. The detector PRF is modelled by multiplying the resulting \ac{psf} by a pixel sensitivity map which is drawn from a normal distribution with mean 0 and standard deviation 0.05. The histogram of pixel sensitivities and total pixel sensitivity map is shown in the right panel of Figure~\ref{fig:optics}.

This operation is then performed over 20 stars with a uniform 100nm bandpass spectra sampled at three wavelengths. This set of stars is then imaged four times with telescope pointing dithers applied between each image. This is done to break the natural degeneracy of stellar fluxes and pixel sensitivities. This final model complexity requires the calculations of 240 individual \ac{psf}s over a 1024x1024 detector, which takes a total of \textcolor{red}{showyourwork variable} seconds to evaluate on an Apple M1 CPU. The data set is generated by taking this set of images and applying Poissonian noise to the \ac{psf}s and additive Gaussian noise to every pixel as detector read-noise. Using this set of images we recover stellar positions and fluxes, optical aberrations, and individual pixel sensitivities. This gives us 40 stellar positional parameters (20 RA and DEC each), 20 stellar flux parameters, 7 optical aberration parameters, and 1'048'576 detector parameters. The dimensionality is vastly dominated by the \ac{prf}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/data_resid.pdf}
    \caption{The left panel shows a single image of the full four that is used to create the full data set. This image has had both photon and detecotor noise applied. By eye it is clearly difficult to detangle the astrophysical information. The large number of overlapping \ac{psf}s is chosen in order to spread light across the majority of the detector so that we encode the PRF information for as much of the detector as possible. The middle panel shows the residual of this single image from the data and the initial uncalibrated model. Clearly these residuals are large, showing that there is a large amount of calibration required. The right panel shows this same residual after the model has been optimised. The residual values are much smaller and at the noise floor. A small zoomed region is shown so that the individual pixel-level residuals can be seen.}
    \label{fig:data_residual}
\end{figure*}

To recover these parameters we initialise a naive unaberrated optical model with a uniform \ac{prf}. Stellar positions and fluxes are similarly perturbed from their true values. We will use this model to recover the true values. Figure~\ref{fig:data_residual} shows the resultant a sample residual of this model with the data.

@@

\textcolor{green}{Only by employing gradient evaluations using autodiff makes this problem computationally tractable. In comparison to the standard finite-differencing gradient evaluation algorithms with a complexity that scales linearly with the number of parameters, a problem of this dimensionality would take of order a million times longer to compute, which would remain impossible regardless of the hardware, let alone laptop CPU. The optimisation is performed for a total of 200 epochs, with each gradient calculation taking \textcolor{red}{showyourwork variable} seconds to evaluate on an Apple M1 Max CPU, leading to total optimisation time of \textcolor{red}{showyourwork variable} minutes.}

\textcolor{green}{The optical propagation algorithm employs matrix Fourier transforms which uses a matrix multiplication to model the propagation between conjugate pupil-focal planes, and so can be greatly improved if deployed onto GPU hardware. Similarly as a test case, we initialise from a very naive state far from the truth, requiring a larger number of optimisation epochs than would be required in a real-world example. Furthermore as a calibration algorithm across a whole detector as an example the complexity of the model is greatly increased, applying this to real data would likely be focused on small detector regions resulting in less computationally intensive models.}

\textcolor{green}{Should this green text be moved to discussion?}

This optimisation strategy performs well, with all of the parameters being recovered down to the noise level. This can be seen in Figure~\ref{fig:data_residual} with the right panel showing a sample residual of the data and model after optimisation, which is clearly only noise. Correlations and residual plots of the positions and fluxes after the optimisation are shown in Figure~\ref{fig:astro_params}, and as expected from the residuals all of these parameters are well recovered and tightly constrained around their true values. Figure~\ref{fig:aberrations} shows the recovery of the optical aberrations, with a final residual aberration of \textcolor{red}{showyourwork variable} nm RMS. Finally, Figure~\ref{fig:flat_field} shows a correlation plot of the true and recovered sensitivity values and the corresponding residual histogram. The points of the correlation plot are colored-coded with the total counts for each pixel, showing as expected that pixels with higher photon counts having a better recovery of the PRF since these pixels have a greater \ac{snr}. Overall the parameters are very well recovered down to the noise level.

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figs/astro_params.pdf}
    \caption{The left panel shows the recovery of the individual RA-Dec positions of each star, with the top section showing the correlation between the true and recovered values in units of arcseconds. The bottom section shows the individual residuals for each. Clearly these parameters have been well recovered through the optimisation. The right panel shows the recovery of the flux parameters in units of photons. The top section shows the correlation between the true and recovered values, and the bottom showing the resulting residuals. \textcolor{red}{Should we address the slight overestimation of fluxes due to the PRF values being a relative measure so a small global covaraince is present?}}
    \label{fig:astro_params}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/aberrations.pdf}
    \caption{This figure shows the recovery of the optical aberrations after optimisation. The left panel shows the amplitudes of the Zernike polynomial coefficients, with the top section showing the correlation plot of the true and recovered values, and the bottom showing the individual residuals. The middle and right panels show the the true total \ac{opd} and the residuals generated by these values respectively.}
    \label{fig:aberrations}
\end{figure*}

>>>>

% Define document class
% \documentclass[twocolumn]{spie}
\documentclass[]{spieman}
% \documentclass[]{spie}
% \documentclass[twocolumn]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage{xspace}
%\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations

\renewcommand{\baselinestretch}{1.0} % Change to 1.65 for double spacing
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{relsize}
\usepackage{acro}
% \usepackage{authblk}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\acsetup{
  single    = false,
  list/sort  = true,
  cite/group = true,
  cite/group/cmd = \cite,
  cite/group/pre = {},
}

@@

\author[a,*]{Louis Desdoigts}
\author[b]{Benjamin J. S. Pope}
\author[a]{Peter G. Tuthill}

@@

% Begin!
% \authorinfo{Send correspondence to L. C. D.: \\E-mail: \href{mailto:louis.desdoigts@sydney.edu.au}{louis.desdoigts@sydney.edu.au}}

@@

% Include email contact information for corresponding author
{\noindent \footnotesize\textbf{Send correspondence to L. Desdoigts: \\E-mail: \linkable{louis.desdoigts@sydney.edu.au }}}

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, by taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework \equinox \cite{kidger2021equinox}, with an optical system consisting of an \equinox module wrapping a stack of `layers' which are themselves \equinox modules applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

@@

\begin{enumerate}
    \item Wavefront initialisation
    \item Application of the pupil-support
    \item Normalisation of the wavefront
    \item Application of the pupil phase mask
    \item Application of the optical aberrations
    \item Propagation to the focal plane
\end{enumerate}

@@

Next the wavefront encounters the binary phase plate diffractive pupil from Toliman \cite{tuthill2018}, with a half-wave total \ac{opd} between the binary regions. We use this as an information-preserving way to spread the \ac{psf} across a large number of pixels which encodes the PRF information for a large number of pixels simultaneously. Alternative methods such as defocusing or using a diffuser \cite{Stefansson2017} come at the cost of eliminating Fourier information which constrain the wavefront and positions. An important consideration which is alleviated by the Toliman pupil is that focal-plane wavefront sensing suffers a sign ambiguity for all phase modes that are inversion-symmetric about the origin, and for unambiguous phase retrieval we therefore need an asymmetric or odd-mode symmetric pupil \cite{Martinache2013}. This pupil and corresponding aberrated \ac{psf} are shown in the left panel of Figure \ref{fig:optics}.

The wavefront then encounters low-order phase errors parameterised by coefficients on Zernike polynomials. We include the first 4 radial terms, ignoring global piston, tip and tilt for a total of 7 terms. Coefficients were randomly drawn from a normal distribution of mean 0 and standard deviation 20\,nm, for a total of \textcolor{red}{showyourwork variance}\,nm RMS total \ac{opd}. These aberrations applied to the pupil mask and resulting \ac{psf} are shown in the middle panel of Figure~\ref{fig:optics}, where we see this induces a large amount of distortion in the \ac{psf}.

This is then propagated to the $1024\times1024$ pixel focal plane by a two-sided matrix Fourier transform \cite{Soummer2007,Martinache2020}. The detector PRF is modelled by multiplying the resulting \ac{psf} by a pixel sensitivity map which is drawn from a normal distribution with mean 0 and standard deviation 0.05. The histogram of pixel sensitivities and total pixel sensitivity map is shown in the right panel of Figure~\ref{fig:optics}.

This operation is then performed over 20 stars with uniform 100\,nm bandpass spectra sampled at three wavelengths. This set of stars is then imaged four times with telescope pointing dithers applied between each image. This is done to break the natural degeneracy of stellar fluxes and pixel sensitivities. This total model requires the calculation of 240 individual \ac{psf}s over a $1024\times1024$ detector, which takes a total of \textcolor{red}{showyourwork variable} seconds to evaluate on an Apple M1 CPU. The data set is generated by taking this set of images and applying Poisson photon noise to the \ac{psf}s and additive Gaussian noise to every pixel as detector read-noise. Using this set of images we recover stellar positions and fluxes, optical aberrations, and individual pixel sensitivities. This gives us 40 stellar positional parameters (20 RA and DEC each), 20 stellar flux parameters, 7 optical aberration parameters, and 1,048,576 detector parameters: the dimensionality of the problem is vastly dominated by the \ac{prf} parameters, which is why this problem is made tractable only by using autodiff.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/data_resid.pdf}
    \caption{Left: single image of the full four that is used to create the full data set. This image has had both photon and detector noise applied. By eye it is clearly difficult to disentangle the astrophysical information. The large number of overlapping \ac{psf}s is chosen in order to spread light across the majority of the detector so that we encode the PRF information for as much of the detector as possible. Middle: the residual of this single image from the data and the initial uncalibrated model. Clearly these residuals are large, showing that there is a large amount of calibration required. Right: this same residual after the model has been optimised. The residual values are much smaller and at the noise floor. A small zoomed region is shown so that the individual pixel-level residuals can be seen.}
    \label{fig:data_residual}
\end{figure*}

To recover these parameters we initialise a naive unaberrated optical model with a uniform \ac{prf}. Stellar positions and fluxes are similarly perturbed from their true values. We will use this model to recover the true values. Figure~\ref{fig:data_residual} shows maps of the initial guess and initial and final residuals.

@@

The optimisation is performed for a total of 200 epochs, with each gradient calculation taking \textcolor{red}{showyourwork variable} seconds to evaluate on an Apple M1 Max CPU, leading to a total optimisation time of \textcolor{red}{showyourwork variable} minutes.

This optimisation strategy performs well, with all of the parameters being recovered down to the noise level. This can be seen in Figure~\ref{fig:data_residual} with the right panel showing a sample residual of the data and model after optimisation, which is clearly only noise. Correlations and residual plots of the positions and fluxes after the optimisation are shown in Figure~\ref{fig:astro_params}, and as expected from the residuals all of these parameters are well recovered and tightly constrained around their true values. Figure~\ref{fig:aberrations} shows the recovery of the optical aberrations, with a final residual aberration of \textcolor{red}{showyourwork variable} nm RMS. Finally, Figure~\ref{fig:flat_field} shows a correlation plot of the true and recovered sensitivity values and the corresponding residual histogram. The points of the correlation plot are colour-coded with the total counts for each pixel, showing as expected that pixels with higher photon counts having a better recovery of the PRF since these pixels have a greater \ac{snr}. Overall the parameters are very well recovered down to the noise level.

The optical propagation algorithm employs two-sided matrix Fourier transforms \cite{Soummer2007}, which can be greatly improved if deployed onto GPU hardware. In our a test case, we initialise from a very naive state far from the truth, requiring a larger number of optimisation epochs than would be required in a real-world example. Furthermore as a calibration algorithm across a whole detector as an example the complexity of the model is greatly increased, applying this to real data would likely be focused on small detector regions resulting in less computationally intensive models.

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figs/astro_params.pdf}
    \caption{Left: recovery of the individual RA-Dec positions of each star, with the top section showing the correlation between the true and recovered values in units of arcseconds. The bottom section shows the individual residuals for each. Clearly these parameters have been well recovered through the optimisation. Right: the recovery of the flux parameters in units of photons. The top section shows the correlation between the true and recovered values, and the bottom showing the resulting residuals. \textcolor{red}{Should we address the slight overestimation of fluxes due to the PRF values being a relative measure so a small global covaraince is present?}}
    \label{fig:astro_params}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/aberrations.pdf}
    \caption{Recovery of the optical aberrations after optimisation. Left: the amplitudes of the Zernike polynomial coefficients, with the top section showing the correlation plot of the true and recovered values, and the bottom showing the individual residuals. Middle and right: the true total \ac{opd} and the residuals generated by these values respectively.}
    \label{fig:aberrations}
\end{figure*}
